---
title: "visualizations-page"
---
## Describing data

1.  The first data set is from Cal Recycle and is the Multi-Year County Wide Origin Summary for Monterey. It is found at <https://www2.calrecycle.ca.gov/LGCentral/DisposalReporting/Origin/CountywideSummary> on Cal Recycle's website. This data is called `CountywideSummary.xlsx` The county wide disposal summary for Monterey county contains the variables Report Year; Report Quarter, which lists each of the four quarters of the year and is broken down in to cities with in Monterey county; Disposal Ton, the amount of waste produced within a city within that time period; Export Ton; Transformation Ton; and Total ADC, ADC stands for alternate daily cover which is material other than soil used to cover a landfill.

    The second data source (broken in to two tables) is from the US Census Bureau QuickFacts table. It contains some estimates from vintage year 2024 and 2025 other years date back to 2010. The Census data is from the 2010 Census of Population and Housing and the 2020 Census of Population and Housing. Other data comes from a variety of sources. All data was accessed on March 22, 2026. It can be found at <https://www.census.gov/quickfacts/fact/table/pacificgrovecitycalifornia,montereycitycalifornia,marinacitycalifornia,kingcitynorthcarolina,greenfieldcitycalifornia,gonzalescitycalifornia/PST045224> and <https://www.census.gov/quickfacts/fact/table/soledadcitycalifornia,seasidecitycalifornia,salinascitycalifornia/PST045224>. This data is called `QuickFacts Mar-22-2026 (1).csv` and `QuickFacts Mar-22-2026.csv` The data set contains a variety of infromation on people, businesses and geography of cities over 5,000 people in Monterey County, CA. The variables are Fact, which contains the name of the statistic listed; Fact Note; city name for each city, which contains the value for the respective fact; and Value Note for each city.

## Data Cleaning

The population data sets were combined and the orientation switched to be easier to work with. The city names were simplified and column types were changed. For the waste data set missing values for year were fixed as one year corresponded to several rows. Information on quarter was extracted from the Report Quarter variable and values for quarter and yearly totals were dropped. Quarter was combined with year to allow for a continuous time scale, such that each quarter is displayed as the start of that fourth of the year (ex. quarter 2, 1995 becomes 1995.25). The two data sets were combined and only census data for 2010 was kept for city population and population density per square mile. 2010 data was chosen over 2020 data because waste data only goes to 2019. Waste generation per person was calculated by dividing the waste generation for a city by the population in 2010 of that city.

## Visualizations

```{r}
#| label: setup
#| include: false
#loading all required packages
library(tidyverse)
```

```{r}
#| label: reading-data
#| include: false
#reading in all of the cleaned data
waste_population_q1_q2 <-read_csv("visualizations-page/data-clean/waste-population-q1-q2")
waste_summary_q3 <- read_csv("visualizations-page/data-clean/waste-summary-q3")
```

### How has waste generation changed from 1995 in Monterey County?

```{r}
#| label: visualization-q3
#making a visualization to answer research question 3
#storing plot as an object to allow for different looking graphs for different color schemes

visualization_q3 <- waste_summary_q3 |>
ggplot(aes(x = time,
                y = disposal_ton, 
                color = fct_reorder2(city, time, disposal_ton))
       ) +
  geom_line() +
  labs(title = "Waste Generation in Monterey County by City (1995-2019)",
       x = "Year",
       y = "Disposal Amount per Quarter (Tons)",
       color = "City",
       caption = "Waste disposal amounts in Monterey County for each quarter from 1995 to 2019.
Data from CalRecycle",
       alt = "The graph displays waste generation in Monterey County by city from 1995 to 2019. The line graph has year on the x axis and disposal amount in tons on the y axis, the lines are grouped by city. The lines are variable, often spiking in unision yearly. The overall trend shows a dip from around 2010 to 2015. Data from CalRecycle."
       )
```

::: light-content
```{r}
#| label: visualization-q3-light
#this will display in light mode
#https://quarto.org/docs/output-formats/html-themes.html for info on light and dark modes

visualization_q3 +
  scale_color_viridis_d() +
  theme_bw()

```
:::

::: dark-content
```{r}
#| label: visualization-q3-dark
#this will display in dark mode

visualization_q3 +
  scale_colour_viridis_d() +
  theme_dark(paper = "black", ink = "white")
```
:::

The graph shows that waste generation has varied overtime with a high level of variability each year. There seems to be a dip from 2010 to 2015, however this could be due to the end of data on a number of cities.

```{r}
#| label: start-salinas-valley-solid-waste-authority
#| eval: false
#| include: false
#figuring out exactly when Salinas Valley Solid Waste Authority reporting starts and cities end. For interpreting q3 graph and future graphs.
waste_summary_q3 |>
  filter(city == "Salinas Valley Solid Waste Authority") |>
  arrange(city)

waste_summary_q3 |>
filter(time == 2012.75)

waste_summary_q3 |>
filter(time == 2013)
```

In 2013 data for Gonzales, Greenfeild, King City, Salinas, and Soledad is no longer reported and data for the Salinas Valley Solid Waste Authority begins.

The Salinas Valley Solid Waste Authority operates in Greenfeild, Gonzales, King City, Salinas, Soledad, and part of unincorporated Monterey County. (Salinas Valley Solid Waste Authority) <https://svswa.org/government/formation-service-area/>

### How does population size of a city, for cities over 5000 people, affect waste generation per person in Monterey county?

```{r}
#| label: visualization-q1
#chose to filter out data after 2013 because some cities stop reporting in 2013

visualization_q1 <- waste_population_q1_q2 |>
  filter(time < 2013) |>
  ggplot(aes(x = population_census_april_1_2010,
             y = disposal_per_population_ton)
         ) +
  geom_point(aes(color = fct_reorder(city, population_census_april_1_2010)),
             position = position_jitter(height = 0, width = 400)
             ) +
  geom_smooth(method = lm,
              color = "pink3"
              ) +
  labs(title = "Waste Generation per Person based on Population of Cities (1995-2013)",
       x = "Population (April 1, 2010)",
       y = "Waste Disposal per Person per Quarter (Tons)",
       color = "City",
       caption = "Waste generation per person for cities over 5000 people in Monterey County based on population size 
       from 1995 up to 2013.
       Data from US Census Bureau and CalRecycle",
       alt = "Graph of Waste generation per person per quarter in tons for cities over 5000 people in Monterey County based on population size, from 1995 up to 2013. Population is based on the 2010 census, April 1, 2010. The graph shows a slight increase in waste generation per person as population size increases. Salinas has the largest population with more than 10000 more people than Seaside the next largest city. The lowest population is Gonzales. Monterey has generally high waste generation per person. King City has several very low points. Data from US Census Bureau and CalRecycle"
       )
```

::: {.light-content}
```{r}
#| label: visualization-q1-light
# displaying in light mode
visualization_q1 +
  scale_colour_viridis_d() +
  theme_bw()
```
:::

::: {.dark-content}
```{r}
#| label: visualization-q1-dark
#this will display in dark mode

visualization_q1 +
  scale_colour_viridis_d() +
  theme_dark(paper = "black", ink = "white")
```

:::

Because data for some cities stops and is grouped into The Salinas Valley Solid Waste Authority after 2013 the graph only includes data up to 2013. The box plot shows a slight increase in waste generation per person as population size increases. Population size does not seem to be the only factor affecting waste generation, although there is a slight increase in waste generation per person as population increases.

### How does population size per square mile affect waste generation per person for cities over 5000 people in Monterey County?

```{r}
#| label: visualization-q2
#choosing not to include data from after 2013 because Salinas Valley Solid Waste Authority data begins

visualization_q2 <- waste_population_q1_q2 |>
  filter(time < 2013) |>
  ggplot(aes(x = population_per_square_mile_2010,
             y = disposal_per_population_ton)
         ) +
  geom_point(aes(color = fct_reorder(city, population_per_square_mile_2010)),
             position = position_jitter(height = 0, width = 30),
             show.legend = FALSE
             ) +
  geom_smooth(color = "red",
              method = lm
              ) +
  labs(title = "Population Density Versus Waste per Person by City (1995-2013)",
       x = "Population Density per Square Mile (2010)",
       y = "Waste Disposal per Person per Quarter (Tons)",
       caption = "City population density versus waste generation per person for cites over 5000 people in Montery County 
from 1995 up to 2013.
       Data from US Census Bureau and CalRecycle",
       alt = "A graph displaying population density per square mile in 2010 versus waste disposal per person per quarter in tons by city for cities over 5000 people in Monterey County from 1995 to 2013. The graph shows a negative corrolation between waste generation and population density. Data from US Census Bureau and CalRecycle"
      ) +
  annotate("text", x = 2219, y = 0.11, label = "Marina") +
  annotate("text", x = 3285, y = 0.5, label = "Monterey") +
  annotate("text", x = 3348, y = 0.08, label = "King City") +
  annotate("text", x = 3575, y = 0.4, label = "Seaside") +
  annotate("text", x = 4260, y = 0.03, label = "Gonzales") +
  annotate("text", x = 5250, y = 0.35, label = "Pacific Grove") +
  annotate("text", x = 5831, y = 0.01, label = "Soledad") +
  annotate("text", x = 6490, y = 0.35, label = "Salinas") +
  annotate("text", x = 7648, y = 0.01, label = "Greenfeild")
```


::: {.light-content}
```{r}
#| label: visualization-q12-light
# displaying in light mode
visualization_q2 +
  scale_colour_viridis_d() +
  theme_bw()
```
:::

::: {.dark-content}
```{r}
#| label: visualization-q2-dark
#this will display in dark mode

visualization_q2 +
  scale_colour_viridis_d() +
  theme_grey(paper = "black", ink = "white")
```
:::

The graph shows that as population density increases waste disposal per person tends to decrease.

```{r}
#| label: box-plot-for-q2
# classifying by urban and rural

boxplot_q2 <- waste_population_q1_q2 |>
  mutate(rural_urban = case_when(population_per_square_mile_2010 <= 4000 ~ "more rural",
                                 population_per_square_mile_2010 >= 4000 ~ "more urban")) |>
  ggplot() +
  geom_boxplot(aes(y = disposal_per_population_ton,
                   color = rural_urban
                   ),
               show.legend = FALSE, 
               ) +
  labs(title = "City Type Versus Waste Generation per Person (1995-2013)",
       y = "Waste Disposal per Person per Quarter (Tons)",
       color = "City Type",
       caption = "Waste generation per person for more urban versus more rural cities based on population density per 
       square mile as of 2010 for cites over 5000 people in Montery County from 1995 up to 2013.
       Data from US Census Bureau and CalRecycle",
       alt = "The boxplot displays waste generation per person per quarter in tons for more urban versus more rural cities based on population density per square mile as of 2010 for cites over 5000 people in Montery County. Data from 1995 up to 2013. More rural cities have a higher mean waste generation per person and have several outliers. Data from US Census Bureau and CalRecycle."
       ) +
  scale_x_discrete(labels = NULL, continuous.limits = c(-0.001, 0.001)) +
  annotate("text", x = -0.2, y = -0.02, label = "more rural") +
  annotate("text", x = 0.2, y = -0.02, label = "more urban")
```


::: {.light-content}
```{r}
#| label: boxplot-q2-light
# displaying in light mode
boxplot_q2 +
  theme_bw() +
  scale_color_manual(values = c("green3", "purple"))
```
:::

::: {.dark-content}
```{r}
#| label: boxplot-q2-dark
#this will display in dark mode

boxplot_q2 +
  theme_dark(paper = "black", ink = "white") +
  scale_color_manual(values = c("blue", "red"))
```

:::
Cities are categorized as more urban or more rural based on the top 5 population densities are the more urban areas and the bottom 4 population densities are the more rural areas. In general for Monterey County the more urban areas have a lower waste generation per person than the more rural areas.

